---
title: Delta DNA methylation β value regression analysis
editor_options: 
  chunk_output_type: console
---

This script performs probe-wise linear regression comparing changes in DNA methylation (delta β values) from baseline to the end of treatment between the vitamin C and placebo groups.

<hr>

### Required R packages
```{r}
#| label: Required packages
#| warning: false
#| message: false
.libPaths("/packages/")
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  fs,
  data.table,
  Matrix,
  matrixStats,
  stringr,
  purrr,
  doFuture,
  qs
)
```

### Project setup
```{r}
#| label: Analysis settings
#| warning: false
#| message: false
project_dir <- path("~/EVITA_EPIC_methylation_analysis")
r_data_dir <- path(project_dir, "r_data")
# parallel processing settings
n_workers <- parallelly::availableCores() - 2
future::plan(multisession, workers = n_workers)
options(future.globals.maxSize = 10000 * 1024^2)
# Set random seed for reproducibility
set.seed(1022)
```

### DNA methylation data loading
- Delta β value matrix (End of treatment - Baseline) and sample metadata
```{r}
#| label: Load data
fname <- "EVITA_DNA_methylation_EPIC_delta_beta_values"
dt <- readRDS(path(r_data_dir, fname, ext = "rds"))
fname <- "EVITA_DNA_methylation_EPIC_sample_table"
sample_table <- readRDS(path(r_data_dir, fname, ext = "rds"))
sample_table <- sample_table[Timepoint != "Baseline"]
stopifnot(identical(colnames(dt)[-1], sample_table[, PatientID]))
```

### Sample selection (Mutant or wildtype)
Select either wild-type or mutant samples for the regression analysis.
- Select wild-type samples
```{r}
#| label: Sample selection wildtype
group <- "wildtype"
sample_table[TET2 == 0 & IDH == 0][, .N, .(TET2, IDH)]
idat_ids <- sample_table[TET2 == 0 & IDH == 0, IDAT]
```
- Select mutant samples
```{r}
#| label: Sample selection mutant
#group <- "mutant"
#sample_table[TET2 == 1 | IDH == 1][, .N, .(TET2, IDH)]
#idat_ids <- sample_table[TET2 == 1 | IDH == 1, IDAT]
```

### Probe filtering based on missing data
```{r}
#| label: Probe filtering based on missing data
fname <- "EVITA_DNA_methylation_EPIC_probe_metadata"
probe_meta <- readRDS(path(r_data_dir, fname, ext = "rds"))
probe_ids <- probe_meta[percent_missing_vals < 20, ProbeID]
dt <- dt[ProbeID %in% probe_ids]
```

### Regression data matrix
```{r}
#| label: Regression data matrix
sample_table <- sample_table[IDAT %in% idat_ids]
sample_table[, .N, .(TET2, IDH)]
patient_ids <- sample_table[, PatientID]
dt <- dt[, .SD, .SDcols = c("ProbeID", patient_ids)]
df <- as.matrix(dt, rownames = "ProbeID")
identical(sample_table[, PatientID], colnames(df))
probe_ids <- rownames(df)
length(probe_ids)
```

### Treatment labels
- Generate randomised treatment labels
```{r}
#| label: Treatment labels
# Original treatment label
sample_table[, .N, Treatment_arm]
original_label  <- copy(sample_table[, Treatment_arm])

# Treatment label scrambled
n_permutations <- 1000
scrambled_labels <- matrix(nrow = length(original_label), ncol = n_permutations)
for (run in 1:n_permutations) {
  scrambled_labels[, run] <- sample(original_label)
}
dt_scrambled_labels <- as.data.table(scrambled_labels)
setnames(dt_scrambled_labels, str_c("scramble_", 1:n_permutations))
dt_scrambled_labels[1:5, 1:5]
sample_table <- cbind(sample_table, dt_scrambled_labels)
```

### Define function to run probe-wise linear regression for a single treatment label
- Return p-values and regression coefficients for vitamin C vs placebo for each probe
```{r}
#| label: Define function
run_single_regression <- function(treatment_labels) {
  reg_results <- foreach(
    i = seq_along(probe_ids),
    .combine = "rbind"
  ) %dofuture% {
    tryCatch({
      betas <- as.numeric(df[i, ])
      fit <- lm(betas ~ treatment_labels)
      fit <- setDT(broom::tidy(fit))
      data.table(
        b_vit_c = fit[term == "treatment_labelsVitamin C", estimate],
        pval_vit_c = fit[term == "treatment_labelsVitamin C", p.value]
      )
    },
    error = function(e) {
      # Return NA values if there's an error
      message("Error in probe ", i, ": ", e$message)
      data.table(
        b_vit_c = NA_real_,
        pval_vit_c = NA_real_
      )
    },
    warning = function(w) {
      # Log warning but continue processing
      message("Warning in probe ", i, ": ", w$message)
      NULL
    })
  }
  reg_results <- data.table(reg_results)
  reg_results <- cbind(ProbeID = probe_ids, reg_results)
  return(reg_results)
}
```

### Perform probe-wise linear regression for original and scrambled treatment labels
```{r}
#| label: Run regression
# Original label
original_results <- run_single_regression(original_label)
original_results

# Scrambled labels
scrambled_results <- list()
for (i in 1:n_permutations) {
  cat(sprintf("\rProcessing permutation %d/%d", i, n_permutations))
  cat("\n")
  scrambled_label <- sample_table[[str_c("scramble_", i)]]
  scrambled_results[[i]] <- run_single_regression(scrambled_label)
}
cat("\nCompleted all permutations\n")
```

### Save results for later use
```{r}
#| label: Save results
results <- list(
  original = original_results,
  scrambled = scrambled_results
)
fname <- str_c("EVITA_delta_betas_regression_results_", group)
qsave(results, file = path(r_data_dir, fname, ext = "qs"))
```

### R Session
```{r}
#| label: R Session
sessionInfo()
```