---
title: GO term overrepresentation analysis of genepromoter CpGs
editor_options: 
  chunk_output_type: console
---

This script performs a gene ontology (GO) term overrepresentation analysis on the differentially methylated CpGs in gene promoter regions.

<hr>

### Required packages
```{r}
#| label: Setup
#| warning: false
.libPaths("/packages/")
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidyverse,
  data.table,
  fs,
  clusterProfiler,
  org.Hs.eg.db,
  qs
)
path <- fs::path
```

### Project setup
```{r}
#| label: project setup
#| warning: false
project_dir <- path("~/EVITA_EPIC_methylation_analysis")
r_data_dir <- path(project_dir, "r_data")
```

### Regression results based on wild-type samples
- Delta_beta_value_regression.qmd for details
```{r}
#| label: Wild-type regression results
fname <- "EVITA_delta_betas_regression_results_wildtype"
dt_results <- qread(path(r_data_dir, fname, ext = "qs"))
dt <- dt_results$original
dt
```

### Probe annotation and subsetting (gene promoter CpGs)
- EPICv1 probe manifest and probe-gene mapping:
  - https://zhouserver.research.chop.edu/InfiniumAnnotation
- Promoter CpGs are defined as 1500 bp flanking regions upstream and downstream of transcription start sites (TSS)
```{r}
#| label: Probe manifest
fname <- "EPICv1_manifest_hg38"
dt_probe <- readRDS(path(r_data_dir, fname, ext = "rds"))

# Gene promoter annotation
fname <- "EPICv1_manifest_hg38_probe_gene_mapping"
gene_mapping <- readRDS(path(r_data_dir, fname, ext = "rds"))
gene_mapping <- gene_mapping[order(ProbeID, absolute_dist_to_tss)]
gene_mapping <- gene_mapping[!duplicated(Gene_probeID)]
probe_ids <- gene_mapping[absolute_dist_to_tss < 1500, ProbeID]
dt_probe[, tss_1500 := ProbeID %in% probe_ids]
probe_ids_subset <- dt_probe[tss_1500 == TRUE, ProbeID]

# Subset the regression results to promoter CpGs
dt <- dt[ProbeID %in% probe_ids_subset]

# Merge with gene mapping annotation
dt <- merge(dt, gene_mapping, by = "ProbeID")

# Subset to promoter CpGs
dt <- dt[absolute_dist_to_tss < 1500]
dt <- dt[order(ProbeID, absolute_dist_to_tss)][!duplicated(ProbeID)]
dt <- dt[b_vit_c < -0.05 & pval_vit_c < 0.01][order(pval_vit_c)]
```

### Identify Entrez IDs and full gene names
```{r}
#| label: Entrez IDs and full gene names
gene_name <- unique(dt[, GeneName])
annot <- AnnotationDbi::select(org.Hs.eg.db,
                               keys = gene_name,
                               columns = c("GENENAME", "ENTREZID"),
                               keytype = "SYMBOL")
setDT(annot)
setnames(annot, "SYMBOL", "Symbol")
```

### Run GO overrepresentation analysis
```{r}
#| label: GO overrepresentation analysis
ggo <- enrichGO(
  gene = annot[, ENTREZID],
  universe = keys(org.Hs.eg.db),
  keyType = "ENTREZID",
  OrgDb = org.Hs.eg.db,
  ont = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff = 1,
  qvalueCutoff = 1,
  readable = TRUE
)
dt <- data.table(ggo@result)

# Odds Ratio calculation
gene_ratio <- as.data.table(
  str_split(
    dt[, GeneRatio],
    pattern = "/",
    simplify = TRUE
  )
)
bg_ratio <- as.data.table(
  str_split(
    dt[, BgRatio],
    pattern = "/",
    simplify = TRUE
  )
)
setnames(bg_ratio, c("V3", "V4"))
dt_or <- cbind(gene_ratio, bg_ratio)
dt_or <- dt_or[, lapply(.SD, as.numeric)]
dt_or[, V2 := V2 - V1]
dt_or[, V4 := V4 - V3]
dt_or[, OR := (V1 / V2) / (V3 / V4)]

dt <- cbind(dt, dt_or)
dt <- dt[order(p.adjust)]
table <- dt[
  p.adjust < 0.05,
  .(
    ID,
    Description,
    Count,
    round(OR, 2),
    round(p.adjust, 3)
  )
]
setnames(
  table,
  c(
    "Gene_ontology_term",
    "Description",
    "Gene_count",
    "Odds_ratio",
    "FDR_adjusted_Pvalue"
  )
)
```
```{r}
#| label: display_table
#| echo: false
knitr::kable(table, caption = "Enrichment Results")
```

## R Session
```{r}
#| label: R Information
sessionInfo()
```