---
title: Heatmap of delta DNA methylation values
editor_options: 
  chunk_output_type: console
---

This script generates a heatmap displaying the changes in DNA methylation (delta β values) from baseline to the end of treatment.<br>
Gene promoter and distal enhancer CpGs that were hypomethylated in vitamin C-treated patients with wild-type TET2 and IDH1/2 are shown.

<hr>

### Load required R packages
```{r}
#| label: packages
#| warning: false
#| message: false
.libPaths("/packages/")
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  matrixStats,
  tidyverse,
  data.table,
  fs,
  ComplexHeatmap,
  viridis,
  circlize,
  qs
)
```

### Project setup
```{r}
#| label: project setup
#| warning: false
#| message: false
project_dir <- path("~/EVITA_EPIC_methylation_analysis")
r_data_dir <- path(project_dir, "r_data")
# Source color palettes
source(path(project_dir, "EPIC_array_analysis_color_palettes.R"))
# Set seed for reproducibility
set.seed(1022)
```

### DNA methylation data loading
- Delta β value matrix (End of treatment - Baseline) and sample metadata
```{r}
#| label: sample annotation
fname <- "EVITA_DNA_methylation_EPIC_delta_beta_values"
dt <- readRDS(path(r_data_dir, fname, ext = "rds"))
fname <- "EVITA_DNA_methylation_EPIC_sample_table"
sample_table <- readRDS(path(r_data_dir, fname, ext = "rds"))
sample_table <- sample_table[Timepoint != "Baseline"]
stopifnot(identical(colnames(dt)[-1], sample_table[, PatientID]))
```

### Somatic variant data processing
- Select the highest VAF per patient for each gene
```{r}
#| label: Somatic variant data processing
fname <- "EVITA_curated_variants_EPIC_analysis"
vaf <- readRDS(path(r_data_dir, fname, ext = "rds"))
vaf <- vaf[EPIC_Data == TRUE, .(Patient_ID_Timepoint, Hugo_Symbol, AF)]

# TET2
vaf_tet2 <- vaf[Hugo_Symbol == "TET2"]
vaf_tet2 <- vaf_tet2[order(Patient_ID_Timepoint, AF, decreasing = TRUE)]
vaf_tet2 <- vaf_tet2[!duplicated(Patient_ID_Timepoint)]
colnames(vaf_tet2)[-1] <- str_c("TET2_", colnames(vaf_tet2)[-1])

# IDH1/2
vaf_idh1 <- vaf[Hugo_Symbol == "IDH1"]
vaf_idh2 <- vaf[Hugo_Symbol == "IDH2"]
vaf_idhs <- rbind(vaf_idh1, vaf_idh2)
vaf_idhs[, Hugo_Symbol := "IDH"]
colnames(vaf_idhs)[-1] <- str_c("IDH_", colnames(vaf_idhs)[-1])

# DNMT3A
vaf_dnmt3a <- vaf[Hugo_Symbol == "DNMT3A"]
vaf_dnmt3a <- vaf_dnmt3a[order(Patient_ID_Timepoint, AF, decreasing = TRUE)]
vaf_dnmt3a <- vaf_dnmt3a[!duplicated(Patient_ID_Timepoint)]
colnames(vaf_dnmt3a)[-1] <- str_c("DNMT3A_", colnames(vaf_dnmt3a)[-1])

# Merge VAF data
vaf <- merge(vaf_tet2, vaf_idhs, by = c("Patient_ID_Timepoint"), all = TRUE)
vaf <- merge(vaf, vaf_dnmt3a, by = c("Patient_ID_Timepoint"), all = TRUE)

sample_table <- merge(sample_table, vaf, by = "Patient_ID_Timepoint",
                      sort = FALSE, all.x = TRUE)
```

### Regression results for wild-type cases
```{r}
#| label: Regression results
fname <- "EVITA_delta_betas_regression_results_wildtype"
dt_results <- qread(path(r_data_dir, fname, ext = "qs"),
                    nthreads = 8)
dt_results <- dt_results$original
dt_results
```

### Probe annotations and subsetting (gene promoter and distal enhancer CpGs)
- EPICv1 probe manifest and probe-gene mapping:
  - https://zhouserver.research.chop.edu/InfiniumAnnotation
- ENCODE Candidate Cis-Regulatory Elements (cCRE) annotations:
  - https://screen.encodeproject.org/index/cversions
  - Registry V2 (Version used in above-cited Nature paper)
- Promoter CpGs are defined as CpGs with a distance to the TSS less than 1500 bp
```{r}
#| label: Probe manifest
fname <- "EPICv1_manifest_hg38"
dt_probes <- readRDS(path(r_data_dir, fname, ext = "rds"))
fname <- "EVITA_DNA_methylation_EPIC_probe_metadata"
probe_meta <- readRDS(path(r_data_dir, fname, ext = "rds"))

# CpG islands annotation
dt_probes[, cgi := CGIposition %in% "Island"]

# ENCODE distal enhancer annotation
probe_ids <- probe_meta[cCRE_ENCODE %like% "dELS", ProbeID]
dt_probes[, dist_enh := ProbeID %in% probe_ids]

# Gene promoter annotation
fname <- "EPICv1_manifest_hg38_probe_gene_mapping"
gene_mapping <- readRDS(path(r_data_dir, fname, ext = "rds"))
gene_mapping <- gene_mapping[order(ProbeID, absolute_dist_to_tss)]
gene_mapping <- gene_mapping[!duplicated(Gene_probeID)]
probe_ids <- gene_mapping[absolute_dist_to_tss < 1500, ProbeID]
dt_probes[, tss_1500_cpgs := ProbeID %in% probe_ids]

# Subset CpGs that are in gene promoter region or distal enhancer region
probe_ids_subset <- dt_probes[tss_1500_cpgs == TRUE | dist_enh == TRUE, ProbeID]
```

### Significant CpG selection
- Select CpGs that were hypomethylated in vitamin C-treated patients with wild-type TET2 and IDH1/2
```{r}
#| label: Significant CpG selection
dt_results <- dt_results[ProbeID %in% probe_ids_subset]
dt_results[pval_vit_c < 0.01 & b_vit_c < -0.05][order(-b_vit_c)]
probe_ids <- dt_results[
  pval_vit_c < 0.01 & b_vit_c < -0.05
][
  order(pval_vit_c), ProbeID
]
```

### Heatmap data preparation
- Clip the delta β values to between -0.3 and 0.3
```{r}
#| label: Heatmap data preparation
dt <- dt[ProbeID %in% probe_ids]
df <- as.matrix(dt, rownames = "ProbeID")
df <- pmax(pmin(df, 0.3), -0.3)
dim(df)
stopifnot(identical(colnames(df), sample_table[, PatientID]))
```

### Determine sample order for heatmap
```{r}
#| label: Sample order
sample_table[, TET2_IDH := ifelse(TET2 == 1 | IDH == 1, "Mutant", "Wildtype")]
sample_table[, .N, TET2_IDH]
sample_table[, order_mut := 1]
sample_table[TET2_IDH == "Wildtype", order_mut := 2]
sample_table[, order_treatment := 1]
sample_table[Treatment_arm == "Vitamin C", order_treatment := 2]
sample_table <- sample_table[order(order_mut, order_treatment, PatientID)]
patient_ids <- sample_table[, PatientID]
df <- df[, patient_ids]
stopifnot(identical(colnames(df), sample_table[, PatientID]))
```

### Heatmap split by mutation status and treatment arm
```{r}
#| label: Heatmap split by mutation status and by treatment arm
mut_idx <- sample_table[, TET2_IDH] == "Mutant"
wt_idx <- sample_table[, TET2_IDH] == "Wildtype"

df_mut <- df[, mut_idx]
df_wt <- df[, wt_idx]

sample_table_mut <- sample_table[mut_idx]
sample_table_wt <- sample_table[wt_idx]

# Heatmap sample annotations
column_ha_mut <- HeatmapAnnotation(
  Sex = sample_table_mut[, Sex],
  Age = sample_table_mut[, Age_inclusion_years],
  Diagnosis = sample_table_mut[, Diagnosis],
  DNMT3A_VAF = sample_table_mut[, DNMT3A_AF],
  IDH_VAF = sample_table_mut[, IDH_AF],
  TET2_VAF = sample_table_mut[, TET2_AF],
  Treatment_Arm = sample_table_mut[, Treatment_arm],
  col = list(
    Sex = sex_colors,
    Age = col_fun_age,
    Diagnosis = diagnosis_colors,
    DNMT3A_VAF = col_fun_turbo100,
    IDH_VAF = col_fun_turbo100,
    TET2_VAF = col_fun_turbo100,
    Treatment_Arm = treatment_arm_colors
  ),
  show_annotation_name = FALSE,
  show_legend = FALSE,
  gp = gpar(col = "black", lwd = 0),
  simple_anno_size = unit(0.4, "cm")
)

column_ha_wt <- HeatmapAnnotation(
  Sex = sample_table_wt[, Sex],
  Age = sample_table_wt[, Age_inclusion_years],
  Diagnosis = sample_table_wt[, Diagnosis],
  "DNMT3A VAF" = sample_table_wt[, DNMT3A_AF],
  "IDH VAF" = sample_table_wt[, IDH_AF],
  "TET2 VAF" = sample_table_wt[, TET2_AF],
  Treatment = sample_table_wt[, Treatment_arm],
  col = list(
    Sex = sex_colors,
    Age = col_fun_age,
    Diagnosis = diagnosis_colors,
    "DNMT3A VAF" = col_fun_turbo100,
    "IDH VAF" = col_fun_turbo100,
    "TET2 VAF" = col_fun_turbo100,
    Treatment = treatment_arm_colors
  ),
  show_annotation_name = TRUE,
  show_legend = FALSE,
  gp = gpar(col = "black", lwd = 0),
  simple_anno_size = unit(0.4, "cm")
)

# Heatmap probe annotations
probe_ids <- rownames(df_mut)
length(probe_ids)
dt_probes <- dt_probes[ProbeID %in% probe_ids]
stopifnot(identical(dt_probes[, ProbeID], rownames(df_mut)))
row_ha_binary <- rowAnnotation(
  "CpG Island" = dt_probes[, cgi],
  Promoter = dt_probes[, tss_1500_cpgs],
  Enhancer = dt_probes[, dist_enh],
  col = list(
    "CpG Island" = binary_colors,
    Promoter = binary_colors,
    Enhancer = binary_colors
  ),
  show_legend = FALSE,
  show_annotation_name = TRUE,
  annotation_name_side = "top"
)

# Heatmaps
ht_mut <- Heatmap(df_mut,
  cluster_columns = FALSE,
  show_column_dend = FALSE,
  show_column_names = FALSE,
  column_split = sample_table_mut[, Treatment_arm],
  column_gap = unit(1, "mm"),
  cluster_column_slices = FALSE,
  column_names_rot = 45,
  cluster_rows = TRUE,
  show_row_dend = FALSE,
  show_row_names = FALSE,
  col = col_fun_delta,
  top_annotation = column_ha_mut,
  left_annotation = row_ha_binary,
  show_heatmap_legend = FALSE,
  column_title = "Mutant",
  column_title_gp = gpar(fontsize = 12, fontface = "bold"),
  use_raster = FALSE,
  border = TRUE,
  border_gp = gpar(lwd = 0.2)
)

ht_wt <- Heatmap(df_wt,
  cluster_columns = FALSE,
  show_column_dend = FALSE,
  show_column_names = FALSE,
  column_split = sample_table_wt[, Treatment_arm],
  column_gap = unit(1, "mm"),
  cluster_column_slices = FALSE,
  column_names_rot = 45,
  cluster_rows = TRUE,
  show_row_dend = FALSE,
  show_row_names = FALSE,
  col = col_fun_delta,
  top_annotation = column_ha_wt,
  show_heatmap_legend = TRUE,
  heatmap_legend_param = list(title = "Delta β\nvalue"),
  column_title = "Wildtype",
  column_title_gp = gpar(fontsize = 12, fontface = "bold"),
  use_raster = FALSE,
  border = TRUE,
  border_gp = gpar(lwd = 0.2)
)
```
```{r}
#| label: heatmap plot
#| fig-align: left
#| dpi: 300
#| fig-format: png
#| out-width: "70%"
#| out-height: "70%"
ht_list <- ht_mut + ht_wt
ht_drawn <- draw(ht_list, gap = unit(0.5, "cm"))
```

### Heatmap sample color legends
```{r}
#| label: heatmap sample color legends
column_legends <- packLegend(
  Legend(title = "Sex",
         at = names(sex_colors),
         legend_gp = gpar(fill = sex_colors)),

  Legend(title = "Age",
         col_fun = col_fun_age,
         title_gp = gpar(fontsize = 10),
         direction = "horizontal"),

  Legend(title = "Diagnosis",
         at = names(diagnosis_colors),
         legend_gp = gpar(fill = diagnosis_colors)),

  Legend(title = "VAF",
         col_fun = col_fun_turbo100,
         title_gp = gpar(fontsize = 10),
         direction = "horizontal"),

  Legend(title = "Treatment",
         at = names(treatment_arm_colors),
         legend_gp = gpar(fill = treatment_arm_colors))
)
```
```{r}
#| label: color legends
#| fig-align: left
#| dpi: 300
#| fig-format: png
#| out-width: "50%"
#| out-height: "50%"
draw(column_legends)
```

### R Session
```{r}
#| label: R Information
sessionInfo()
```