---
title: Genomic feature overrepresentation analysis of distal enhancer CpGs
editor_options: 
  chunk_output_type: console
---

This script performs the genomic feature overrepresentation analysis of the differentially methylated CpGs in distal enhancer regions.

<hr>

### Load required R packages
```{r}
#| label: packages
#| warning: false
#| message: false
.libPaths("/packages/")
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidyverse,
  data.table,
  fs,
  sesame,
  qs
)
path <- fs::path
n_threads <- parallel::detectCores() - 2
n_threads
```

### Project directory and data directory
```{r}
#| label: project setup
#| warning: false
#| message: false
project_dir <- path("~/EVITA_EPIC_methylation_analysis")
r_data_dir <- path(project_dir, "r_data")
```

### Regression results based on wild-type samples
- Delta_beta_value_regression.qmd for details
```{r}
#| label: Wild-type regression results
fname <- "EVITA_delta_betas_regression_results_wildtype"
dt_results <- qread(path(r_data_dir, fname, ext = "qs"),
                    nthreads = n_threads)
dt <- dt_results$original
dt
```

### Probe annotation and distal enhancer CpG selection
- EPICv1 probe manifest:
  - https://zhouserver.research.chop.edu/InfiniumAnnotation
- ENCODE Candidate Cis-Regulatory Elements (cCRE) annotations:
  - https://screen.encodeproject.org/index/cversions
  - Registry V2 (Version used in above-cited Nature paper)
```{r}
#| label: Probe manifest
fname <- "EPICv1_manifest_hg38"
dt_probe <- readRDS(path(r_data_dir, fname, ext = "rds"))
fname <- "EVITA_DNA_methylation_EPIC_probe_metadata"
probe_meta <- readRDS(path(r_data_dir, fname, ext = "rds"))
probe_ids <- probe_meta[cCRE_ENCODE %like% "dELS", ProbeID]
dt_probe[, distal_enh := ProbeID %in% probe_ids]

# Use as the universe set for the overrepresentation analysis
probe_ids_subset <- dt_probe[distal_enh == TRUE, ProbeID]
```

### CpG overrepresentation analysis of distal enhancer CpGs
```{r}
#| label: CpG overrepresentation analysis
# Subset to distal enhancer CpGs
dt <- dt[ProbeID %in% probe_ids_subset]
# CpGs that are significantly hypomethylated
probe_ids <- dt[
  pval_vit_c < 0.01 & b_vit_c < -0.05
][
  order(pval_vit_c), ProbeID
]
length(probe_ids)
table <- testEnrichment(query = probe_ids,
                        universe = probe_ids_subset,
                        platform = "EPIC",
                        silent = TRUE)
table <- data.table(table)
table <- table[group %in% c("KYCG.EPIC.TFBSconsensus.20211013",
                            "KYCG.EPIC.tissueSignature.20211211")]
table <- table[order(FDR, p.value)]
table <- table[, .SD,
               .SDcols = c("dbname",
                           "overlap",
                           "estimate",
                           "p.value",
                           "FDR")]
setnames(table,
         c("dbname", "estimate"),
         c("Transcription_factor_or_tissue_signature",
           "Odds_Ratio"))
table <- table[order(FDR, p.value)]
table[, Odds_Ratio := round(Odds_Ratio, 3)]
table[, FDR := round(FDR, 3)]
table[, p.value := formatC(p.value, format = "e", digits = 2)]
table <- table[Odds_Ratio > 1.2 & FDR < 0.01]
```
```{r}
#| label: display_table
#| echo: false
knitr::kable(table, caption = "Results")
```

### R Session
```{r}
#| label: R Information
sessionInfo()
```